Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
working through template proofs

Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented. In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors. If \lstinline{k} is outside the node's range (e.g. because the node has been relocated), the search is restarted from the root node \lstinline{r}.

@@

%% C implementation

When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node. If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findNext} to find the new node. If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).

\subsection{Proofs of Templates} 
In this section, we present our verification methodology, termed \lstinline{traverse}, which applies to both lock-coupling and give-up templates. The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}. Before we delve into the intricacies of these verifications, it is essential to familiarize ourselves with the concept of \emph{atomic triples}.

For many concurrent data structures, the ideal correctness condition requires that the data structure behaves identically to a sequential implementation, even when accessed concurrently by multiple threads. The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}. These take the form of $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple suggests that if $\texttt{P}_l$ holds true before a call and $\texttt{P}_p$ is true for some value of $a$ in a shared state (e.g. a global invariant), then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends). For instance, the specification
$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution. The stack itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the \texttt{is\_stack} assertion.

To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above. This specification assert that the implementation \texttt{push} appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states. The local preconditions and postconditions, \texttt{is\_stack}, contain the points-to and lock-invariant assertions necessary for a thread, while the public preconditions and postconditions, \texttt{stack}, involve the ghost state that defines the overall state of the stack. Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier $g$ (which we will generally omit), it is possible to connect local and public conditions. To show that the implementation of \texttt{push}(v) satisfies these atomic specifications, two things must be demonstrated. Firstly, the implementation can execute safely without owning any piece of public assertion $\texttt{stack}$, only accessing it automatically and restoring it up until the linearization point. Secondly, at some linearization point, the implementation transforms the current stack $s$ into $v::s$ atomically, which satisfies the public postcondition $\texttt{stack}(v::s)$.

\subsubsection{Lock-coupling Template}
\label{traverse_proof_lock}
Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template, as illustrated in Figure \ref{fig:traverse_lock}.
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			 \forall \  \texttt{m}. \left\langle
				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{m} \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow \texttt{(pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}

The specifics of $\nodeboxrep$ and $\treerep$ may vary across implementations, but fundamentally, $\nodeboxrep$ encompasses resources associated with each client thread, while $\treerep$ serves as an abstract representation containing ghost state that links the data structure's concrete state to its abstract equivalent, treated as an atomically accessed shared resource. In particular, $\nodeboxrep(\texttt{pn->n})$ acts as a pointer to both the node's lock and lock-invariant assertions, representing a protected resource through an arbitrary separation logic predicate. Acquiring this lock allows a thread to access additional elements within the data structure.

As previously mentioned, the \lstinline{pn} struct includes two nodes of type \lstinline{node_t}: the $\texttt{p}$ field, signifying the parent node, and the $\texttt{n}$ field, representing the current node. We use the assertion $\ell \lockvar \texttt{R}$ to associate a resource invariant \texttt{R} with a lock location $\ell$. Thus, formally, we can express $\nodeboxrep \texttt{(pn->n)} \triangleq\ \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.

Keep in mind that in the lock-coupling style, each thread consistently retains at least one lock while traversing between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, the thread must acquire a lock for the current node $\texttt{pn->n}$ and obtain the resource safeguarded by the lock. This resource, represented as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the node's concrete information.

>>>>

Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented. In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors. If \lstinline{k} is outside the node's range (e.g. because the node has been relocated), the search is restarted from the root node \lstinline{r}. As in the previous section, we implement this in C with a loop: in each iteration, we acquire the lock on the current node, check that \lstinline{k} is in range, and then use \lstinline{findNext} as above, releasing the lock before we move to the next node. If the \lstinline{inRange} call fails, we release our current lock and return to the root node by setting the current node to the root pointer \lstinline{p} (lines 21-22 in Figure \ref{traverse_giveup_b}).

@@

\subsection{Template Proofs} 
In this section, we show how to verify the templates presented above, by proving that each \lstinline{traverse} function meets its specification. %The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}.
But first, we need to understand how to specify the correctness of \lstinline{traverse} independently of the data structure operations (insert, lookup, etc.) it will be used to implement. The key is the concept of \emph{logical atomicity}~\cite{tada}, a highly compositional approach to specifying concurrent data structure operations. \wm{Should this go in background instead?}

Logical atomicity is a CSL technique for concisely specifying the behavior of a concurrent operation. A logically atomic triple has the form $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread. More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends). For instance, the specification
$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
expresses the fact that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context.

%To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above. This specification assert that the implementation \texttt{push} appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states. The local preconditions and postconditions, \texttt{is\_stack}, contain the points-to and lock-invariant assertions necessary for a thread, while the public preconditions and postconditions, \texttt{stack}, involve the ghost state that defines the overall state of the stack. Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier $g$ (which we will generally omit), it is possible to connect local and public conditions. To show that the implementation of \texttt{push}(v) satisfies these atomic specifications, two things must be demonstrated. Firstly, the implementation can execute safely without owning any piece of public assertion $\texttt{stack}$, only accessing it automatically and restoring it up until the linearization point. Secondly, at some linearization point, the implementation transforms the current stack $s$ into $v::s$ atomically, which satisfies the public postcondition $\texttt{stack}(v::s)$.

Logical atomicity is key to the template approach: each template's \lstinline{traverse} function is proved to satisfy a logically atomic specification that says roughly ``this function finds the node where key \lstinline{k} belongs''. The traverse specifications can then be used to prove atomic specifications for the individual data structure operations, lifting the sequential specifications for insert, lookup, etc. to the concurrent setting.

\subsubsection{Lock-coupling Template}
\label{traverse_proof_lock}

The logically atomic specification for the lock-coupling template's \lstinline{traverse} function is shown in Figure \ref{fig:traverse_lock}. \wm{For the template approach to really work, it seems like all the \lstinline{traverse} functions should have the same specification. Can we make that happen?}
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}

%maybe inline this instead of making it a figure?
\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			 \forall \  \texttt{m}. \left\langle
				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{m} \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow \texttt{(pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}

The definitions of $\nodeboxrep$ and $\treerep$ may vary depending on the data structure, but fundamentally, $\nodeboxrep$ holds the resources owned by each client thread, while $\treerep$ holds the abstract representation of the data structure, treated as an atomically accessed shared resource. In particular, holding $\nodeboxrep(\texttt{pn->n})$ should allow us to acquire the lock on node \texttt{pn->n}, which in turn gives us access to the contents of the node. Formally, we write
\[\nodeboxrep \texttt{(pn->n)} \triangleq\ \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}\]
where the \emph{lock invariant assertion} $\ell \lockvar \texttt{R}$ expresses that the lock $\ell$ protects resources \texttt{R}. In this case, $R$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{pn->n}).

Recall that in the lock-coupling pattern, each thread consistently retains at least one lock while traversing between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, the thread must acquire a lock for the current node $\texttt{pn->n}$ and obtain the resource safeguarded by the lock. This resource, represented as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the node's concrete information.
